### 2017-05-27 paper/conclusions.tex
typo, grammar checked

In this paper, we have proposed an approach to validation of VDM models by discharging proof obligations (POs) with SMT solvers. In our approach, Overture tool is used as the proof obligation generator, and the Python API of Z3 SMT sover is used for encoding and solving formulas of generated POs. Encoding strategies including steps of encoding as well as encoding patterns are demonstrated, and two case studies are reported. From our experiences, it is recognized that discharging POs using SMT solvers is efficient, and the counterexample generation of SMT solvers when the discharging fails can provide useful guidance for correcting inconsistency of VDM models. We also discussed other issues such as applicability of encoding strategies considering the automation of our approach. As the future work, we are gathering and organizing more encoding patterns for challenging the automation issue.

>>>>

In this paper, we have proposed an approach to validation of VDM models by discharging proof obligations (POs) with SMT solvers. In our approach, Overture tool is used as the proof obligation generator, and the Python API of Z3 SMT solver is used for encoding and solving formulas of generated POs. Encoding strategies including steps of encoding as well as encoding patterns are demonstrated, and two case studies are reported. From our experiences, it is recognized that discharging POs using SMT solvers is efficient, and the counterexample generation of SMT solvers when the discharging fails can provide useful guidance for correcting inconsistency of VDM models. We also discussed other issues such as the applicability of encoding strategies considering the automation of our approach. As the future work, we are gathering and organizing more encoding patterns for challenging the automation issue.

### 2017-05-27 paper/introduction.tex
typo, grammar checked

% General information about VDM
The Vienna Development Method (VDM)~\cite{Jones:1990:SSD:94062,Fitzgerald:2005:VDO:1044891,Fitz:PGLarson:2009book} is a formal method which supports modeling and analysis of software systems at various levels of abstraction. A VDM specification can be explicit, implicit, or both. An explicit specification is operational; it describes \emph{how} system components work. An implicit specificiation on the other hand is declarative; it describes \emph{what} system components should work. Software specifications in VDM models define functionalities explicitly or implicitly according to software requirements. VDM has a strong record of applications for design and specification of software systems in industry~\cite{Larsen:2007:RIA:2227886.2227894,DBLP:conf/fm/KuritaCN08,DBLP:journals/ijsi/KuritaN09}. The well-known basic benefits of using VDM to describe software specifications are from the accuracy and unambiguity of VDM, which is common for other formal methods like B.

% Validation methods of VDM
To validate a VDM model, existing tools such as Overture~\cite{Larsen:2010:OII:1668862.1668864} and VDMTools provide static consistency check regarding the syntax and type constraints of VDM specifications. The semantic validation of VDM specifications are established by discharging proof obligations (POs)~\cite{AL:97:POGV} generated by the above tools. POs are logic formulas characterizing consistency of VDM specifications. Theorem provers have been used to discharge the generated POs. Testing (specification animation) by running a VDM model with an interpreter~\cite{Prehn:1991:LNCS551} is the alternative way of validation. This requires that the VDM model to be specified explicitly so that an interpreter can produce specific values for functions and operations.

% More about proof obligations and the motivation: solve as many as possible POs
Discharging POs of VDM models is essential to guarantee the consistency of the formal specifications. For a VDM model, the number of POs is usually large compare to the size of the model. For example, the sample VDM-SL model shown in Fig.~\ref{fig:module_sample} has less than thirty lines, and Overture generates nine POs for the model. Consequently, it is not uncommon to have hundreds of POs for large VDM models. All of them have to be discharged to guarantee consistency of software specifications. Manually discharging so many POs with theorem provers however can be a daunting task in practice. Furthermore, both errorneous VDM specifications and bad proof strategies induce un-dischargeable POs. Theorem provers provide little information when they fail to discharge POs. It is unclear whether a VDM specification is flawed or a better proof strategy is possible when POs cannot be discharged by theorem provers.

@@

In this paper, we apply SMT solvers to discharge POs of VDM specifications. Here we focus on VDM-SL models and the objective is to discharge POs of an VDM-SL model as many as possible through SMT solvers. In our approach, we encode each POs with its context information such as type and state constraints as a first-order formula, and then prove the PO by sending the formula to SMT solvers. The advantages of our approach are (1) the encoding involves only a few segments of a VDM specification; (2) the proof process is automated with SMT solvers; (3) If a PO's proof fails, a counterexample model is returned for further examination.

More specifically, in this approach we chose the Overture tool as the PO generator. For the POs generated by the Overture tool, we encode and prove POs of VDM-SL models by solving SMT formulas with Python API of Z3~\cite{MB:08:ZSS}. Z3 is a popular SMT solver widely used in software verification and its Python API provides flexibility in constructing SMT formulas~\cite{BarFT-SMTLIB}. In order to encode POs, built-in VDM types such as sequences and maps must be characterized by theories in Z3. We conducted case studies and the results showed that our approach can efficiently discharge the significant part of POs of a VDM-SL model.
%  \footnote{POs generated by Overture and VDMTools are slightly different in the numbers and styles of POs}.

The structure of this paper is as follows: Section~\ref{section:vdm} briefly introduce VDM with a sample VDM-SL example; Section~\ref{section:proof-obligations} describes the POs of VDM and explains the POs generated by the Overture tool; section~\ref{section:encode-strategy} describes our strategy of encoding VDM expressions with Python API of Z3; section~\ref{section:case-studies} demonstrates case studies with discussions; section~\ref{section:related-work} compares our approach to related work; section~\ref{section:conclusions} concludes this paper and points out the future work.

>>>>

% General information about VDM
The Vienna Development Method (VDM)~\cite{Jones:1990:SSD:94062,Fitzgerald:2005:VDO:1044891,Fitz:PGLarson:2009book} is a formal method which supports modeling and analysis of software systems at various levels of abstraction. A VDM specification can be explicit, implicit, or both. An explicit specification is operational; it describes \emph{how} system components work. An implicit specification, on the other hand, is declarative; it describes \emph{what} system components should work. Software specifications in VDM models define functionalities explicitly or implicitly according to software requirements. VDM has a strong record of applications for design and specification of software systems in the industry~\cite{Larsen:2007:RIA:2227886.2227894,DBLP:conf/fm/KuritaCN08,DBLP:journals/ijsi/KuritaN09}. The well-known benefits of using VDM to describe software specifications are from the accuracy and unambiguity of VDM, which is common for other formal methods like B.

% Validation methods of VDM
To validate a VDM model, existing tools such as Overture~\cite{Larsen:2010:OII:1668862.1668864} and VDMTools provide static consistency check regarding the syntax and type constraints of VDM specifications. The semantic validation of VDM specifications is established by discharging proof obligations (POs)~\cite{AL:97:POGV} generated by the above tools. POs are logic formulas characterizing consistency of VDM specifications. Theorem provers have been used to discharge the generated POs. Testing (specification animation) by running a VDM model with an interpreter~\cite{Prehn:1991:LNCS551} is the alternative way of validation. This requires that the VDM model to be specified explicitly so that an interpreter can produce specific values for functions and operations.

% More about proof obligations and the motivation: solve as many as possible POs
Discharging POs of VDM models is essential to guarantee the consistency of the formal specifications. For a VDM model, the number of POs is usually large compared to the size of the model. For example, the sample VDM-SL model shown in Fig.~\ref{fig:module_sample} has less than thirty lines, and Overture generates nine POs for the model. Consequently, it is not uncommon to have hundreds of POs for large VDM models. All of them have to be discharged to guarantee consistency of software specifications. Manually discharging so many POs with theorem provers, however, can be a daunting task in practice. Furthermore, both erroneous VDM specifications and bad proof strategies induce un-dischargeable POs. Theorem provers provide little information when they fail to discharge POs. It is unclear whether a VDM specification is flawed or a better proof strategy is possible when POs cannot be discharged by theorem provers.

@@

In this paper, we apply SMT solvers to discharge POs of VDM specifications. Here we focus on VDM-SL models and the objective is to discharge POs of a VDM-SL model as many as possible through SMT solvers. In our approach, we encode each POs with its context information such as type and state constraints as a first-order formula, and then prove the PO by sending the formula to SMT solvers. The advantages of our approach are (1) the encoding involves only a few segments of a VDM specification; (2) the proof process is automated with SMT solvers; (3) If a PO's proof fails, a counterexample model is returned for further examination.

More specifically, in this approach, we chose the Overture tool as the PO generator. For the POs generated by the Overture tool, we encode and prove POs of VDM-SL models by solving SMT formulas with Python API of Z3~\cite{MB:08:ZSS}. Z3 is a popular SMT solver widely used in software verification and its Python API provides flexibility in constructing SMT formulas~\cite{BarFT-SMTLIB}. To encode POs, built-in VDM types such as sequences and maps must be characterized by theories in Z3. We conducted case studies and the results showed that our approach can efficiently discharge the significant part of POs of a VDM-SL model.
%  \footnote{POs generated by Overture and VDMTools are slightly different in the numbers and styles of POs}.

The structure of this paper is as follows: Section~\ref{section:vdm} briefly introduce VDM with a sample VDM-SL example; Section~\ref{section:proof-obligations} describes the POs of VDM and explains the POs generated by the Overture tool; Section~\ref{section:encode-strategy} describes our strategy of encoding VDM expressions with Python API of Z3; Section~\ref{section:case-studies} demonstrates case studies with discussions; Section~\ref{section:related-work} compares our approach to related work; Section~\ref{section:conclusions} concludes this paper and points out the future work.

### 2017-05-27 paper/obligations.tex
typo, grammar checked

% Generation information of POs
For a VDM model, a proof obligation (PO) is a statement that must be proved in order to ensure the consistency of the model. A PO contains a predicate with its context information as shown below. The context information are code segments of the VDM model that relate to the predicate to be proved.

@@

For the sample VDM-SL model shown in Fig.~\ref{fig:module_sample}, Overture generated ten POs. Here selected POs are demonstrated and explained. PO1 is called ``legal sequence application obligation'' that relates to the sequence involved expression (line 9-10) of the VDM-SL model {\tt CMDS}). PO1 says that for all k in the set from 1 to {\tt len s.commands -1}, k should be always within the bound of defined indices of the sequence {\tt s.commands}. Also, it can be recognized that if we use {\tt len s.commands} instead of {\tt len s.commands -1} the proof will fail. Note that in VDM, a sequence are indexed from 1 to its length. Also note that in PO1, the state {\tt CMDS'S} is mentioned, and we have to include the state invariant as well though it is not explicitly described in PO1.

@@

PO7 is an operation postcondition satisfiable obligation for {\tt push\_cmd} (line 15-18). PO7 says that for every input {\tt a:CMD} and {\tt oldstate}, if the precondition whch takes inputs {\tt a} and {\tt oldstate} is satisfied, the postcondition which takes inputs {\tt a}, {\tt oldstate}, and {\tt newstate} should be satisfied. Note that instead of the specification of precondition, postcondition, and state in {\tt CMDS} module, special names {\tt oldstate}, {\tt newstate}, and {\tt post\_push\_cmd} are used. In VDM-SL, the {\tt pre\_} and {\tt post\_} prefix represent the precondition and postcondition of a function/operator separately. They are evaluable functions if given appropriate arguments, and return a boolean value. In a satisfiable obligation of an operation in which the state change is usually expected, the name of state (in this case {\tt commands}) represents {\tt newstate} while the same name postfixed with ``~\textasciitilde~'' (in this case {\tt commands\textasciitilde}) represents {\tt oldstate}.

>>>>

% Generation information of POs
For a VDM model, a proof obligation (PO) is a statement that must be proved to ensure the consistency of the model. A PO contains a predicate with its context information as shown below. The context information is from code segments of the VDM model that relate to the predicate to be proved.

@@

For the sample VDM-SL model shown in Fig.~\ref{fig:module_sample}, Overture generated ten POs. Here selected POs are demonstrated and explained. PO1 is called ``legal sequence application obligation'' that relates to the sequence involved expression (line 9-10) of the VDM-SL model {\tt CMDS}). PO1 says that for all k in the set from 1 to {\tt len s.commands -1}, k should always be within the bound of defined indices of the sequence {\tt s.commands}. Also, it can be recognized that if we use {\tt len s.commands} instead of {\tt len s.commands -1} the proof will fail since a sequence in VDM is indexed from 1 to its length. Also note that in PO1, the state {\tt CMDS'S} is mentioned, and we have to include the state invariant as well though it is not explicitly described in PO1.

@@

PO7 is an operation postcondition satisfiable obligation for {\tt push\_cmd} (line 15-18). PO7 says that for every input {\tt a:CMD} and {\tt oldstate}, if the precondition which takes inputs {\tt a} and {\tt oldstate} is satisfied, the postcondition which takes inputs {\tt a}, {\tt oldstate}, and {\tt newstate} should be satisfied. Note that instead of the specification of precondition, postcondition, and state in {\tt CMDS} module, special names {\tt oldstate}, {\tt newstate}, and {\tt post\_push\_cmd} are used. In VDM-SL, the {\tt pre\_} and {\tt post\_} prefix represent the precondition and postcondition of a function/operator separately. They are evaluable functions if given appropriate arguments, and return a boolean value. In a satisfiable obligation of an operation in which the state change is usually expected, the name of state (in this case {\tt commands}) represents {\tt newstate} while the same name postfixed with ``~\textasciitilde~'' (in this case {\tt commands\textasciitilde}) represents {\tt oldstate}.

### 2017-05-27 paper/related.tex
typo, grammar checked

% Development history of VDM POG.
Proof is essential to guarantee the consistency of VDM models~\cite{978-3-540-19813-0}, and the proof obligation generation of VDM is firstly proposed by B. K. Aichernig and P. G. Larsen~\cite{AL:97:POGV}. Later, A. Ribeiro and P. G. Larsen worked on proof obligation generation and discharging related to termination of recursive functions~\cite{Ribeiro2010}.

% Work related to discharging POs using theorem provers. 
There are several researches focusing on translation from VDM to theorem provers and discharging POs of VDM with theorem provers. S. Agerholm~\cite{Agerholm1996} proposed a translation from a subset of VDM-SL to PVS for type checking. S. Maharaj and J. Bicarregui~\cite{632849} used Agerholm's translation to assist the verification of VDM-SL models and their refinements. Later, S. D. Vermolen et al.~\cite{Verm:2007:master,Vermolen:2010:PCV:1774088.1774608} utilized the parsing mechanism of Overture tools and developed automated translation from a subset of VDM-SL to HOL theorem prover. Vermolen's work also provided some tactics to support automated discharging of POs. L. D. Cuto et al.~\cite{CFP:14:TVCSAP} used the proof obligation generation mechanism of Overture tools for translating CML, a modeling language combing VDM-SL and CSP, to Isabella/HOL for automated proof. Recently, work on improvement of translating VDM-SL to Isabella/HOL are proposed~\cite{CT:15:EOCGTIS}.

% Comparison to our approach
The theorem prover based approaches has common issues: (1) the translation from VDM-SL to theorem provers may be errorneous since the translation applies on the whole VDM-SL model; (2) The proof process is usually complicated and requires an expert even with the help of proof tactics; (3) Theorem provers provide little information when they fail to discharge POs. Compare to our approach that discharges POs of VDM-SL with SMT solvers, we adopt the efficiency of SMT solvers and the encoding does not require to be applied on the whole VDM-SL model. Furthermore, the counterexample generation of SMT solvers is helpful for locating the problem. However, encoding in our approach is manual at this point and still need improvements.

% Work related to validation/verfication of VDM models with model checking.
On the other hand, verification and validation of VDM-SL models with model checking techniques are recently proposed. K. Lausdahl~\cite{kenneth:ifm2013} proposed a semantics-preserving translation that constructs an Alloy model from a subset of VDM-SL. This work aims to support the validation of implicitly specified VDM-SL models with Alloy. H-H. Lin et al.~\cite{DBLP:conf/ftscs/LinOKA15} proposed an approach that combines SPIN model checker and the VDMJ interpreter for verifying explicitly specified VDM-SL models. Lausdahl's work is similar to our approach since Alloy uses SAT solvers, but SMT solvers in our approach can solve wider formulas with more efficiency. Both Lausdahl's and Lin's work require types involved in a VDM-SL model being bounded, and there are issues for Alloy and SPIN to represent the rich and complex VDM-SL types. Compare to them, bounds are not necessary for SMT solvers in our approach and our encoding patterns can applied to complex types thanks for build-in theories in SMT sovers.

>>>>

% Development history of VDM POG.
Proof is essential to guarantee the consistency of VDM models~\cite{978-3-540-19813-0}, and the proof obligation generation of VDM is firstly proposed by B. K. Aichernig and P. G. Larsen~\cite{AL:97:POGV}. Later, A. Ribeiro and P. G. Larsen worked on proof obligation generation and discharging related to the termination of recursive functions~\cite{Ribeiro2010}.

% Work related to discharging POs using theorem provers. 
There are researches focusing on translation from VDM to theorem provers and discharging POs of VDM with theorem provers. S. Agerholm~\cite{Agerholm1996} proposed a translation from a subset of VDM-SL to PVS for type checking. S. Maharaj and J. Bicarregui~\cite{632849} used Agerholm's translation to assist the verification of VDM-SL models and their refinements. Later, S. D. Vermolen et al.~\cite{Verm:2007:master,Vermolen:2010:PCV:1774088.1774608} utilized the parsing mechanism of Overture tools and developed automated translation from a subset of VDM-SL to HOL theorem prover. Vermolen's work also provided some tactics to support automated discharging of POs. L. D. Couto et al.~\cite{CFP:14:TVCSAP} used the proof obligation generation mechanism of Overture tools for translating CML, a modeling language combing VDM-SL and CSP, to Isabella/HOL for automated proof. Recently, work on improvement of translating VDM-SL to Isabella/HOL was proposed~\cite{CT:15:EOCGTIS}.

% Comparison to our approach
The theorem prover based approaches has common issues: (1) the translation from VDM-SL to theorem provers may be erroneous since the translation applies to the whole VDM-SL model; (2) The proof process is usually complicated and requires an expert even with the help of proof tactics; (3) Theorem provers provide little information when they fail to discharge POs. Compare to our approach that discharges POs of VDM-SL with SMT solvers, we adopt the efficiency of SMT solvers and the encoding does not require to be applied on the whole VDM-SL model. Furthermore, the counterexample generation of SMT solvers is helpful for locating the problem. However, encoding in our approach is manual at this point and still need improvements.

% Work related to validation/verfication of VDM models with model checking.
On the other hand, verification and validation of VDM-SL models with model checking techniques are recently proposed. K. Lausdahl~\cite{kenneth:ifm2013} proposed a semantics-preserving translation that constructs an Alloy model from a subset of VDM-SL. This work aims to support the validation of implicitly specified VDM-SL models with Alloy. H-H. Lin et al.~\cite{DBLP:conf/ftscs/LinOKA15} proposed an approach that combines SPIN model checker and the VDMJ interpreter for verifying explicitly specified VDM-SL models. Lausdahl's work is similar to our approach since Alloy uses SAT solvers, but SMT solvers in our approach can solve wider formulas with more efficiency. Both Lausdahl's and Lin's work require types involved in a VDM-SL model being bounded, and there are issues for Alloy and SPIN to represent the rich and complex VDM-SL types. Compare to them, bounds are not necessary for SMT solvers in our approach, and our encoding patterns can be applied to complex types thanks for built-in theories in SMT solvers.
